\ \\ \LARGE\bf Speculative Evaluation in Particle Swarm Optimization

### 2010-04-19 honors_thesis/ppsn_paper/paper.tex
Got the style file for PPSN and made some simple changes to the paper

\documentclass[letterpaper]{sig-alt-release2}

@@

%% PSO Stuff
\DeclareMathOperator{\URand}{U}
\DeclareMathOperator*{\argmin}{arg\;min}
\DeclareMathOperator*{\argmax}{arg\;max}
\DeclareMathOperator*{\arginf}{arg\;inf}
\DeclareMathOperator*{\argsup}{arg\;sup}
\providecommand{\ppos}{\ensuremath{\Vec{x}}}
\providecommand{\pvel}{\ensuremath{\Vec{v}}}
\providecommand{\nbest}{\ensuremath{\Vec{b}_N}}
\providecommand{\pbest}{\ensuremath{\Vec{b}_P}}
\providecommand{\constriction}{\ensuremath{\chi}}
\providecommand{\coeff}{\ensuremath{\phi}}
\providecommand{\obs}{\ensuremath{\Vec{\xi}}}

%SpecExPSO Stuff
\providecommand{\noeval}[1]{\ensuremath{#1^{-e}}}
\providecommand{\nonbest}[1]{\ensuremath{#1^{-n}}}
\providecommand{\p}{\ensuremath{p}}
\providecommand{\pset}{\ensuremath{\mathbf{p}}}
\providecommand{\s}{\ensuremath{s}}
\providecommand{\sset}{\ensuremath{\mathbf{s}}}
\providecommand{\nsset}{\ensuremath{\mathbf{ns}}}
\providecommand{\n}{\ensuremath{n}}
\providecommand{\nset}{\ensuremath{\mathbf{n}}}
\providecommand{\nnset}{\ensuremath{\mathbf{nn}}}

@@

\category{G.1.6}{Numerical Analysis}{Optimization}

\terms{Algorithms}

\keywords{Particle swarm optimization; speculative evaluation; parallel
optimization}

@@

Particle swarm optimization (PSO) has been found to be a highly robust and
effective algorithm for solving many types of optimization problems.  For much
of the algorithm's history, PSO was run serially on a single machine.  However,
the world's computing power is increasingly coming from large clusters of
processors.  Even in desktop machines, systems with 4 or 8 cores are
commonplace.  In order to efficiently utilize these resources for
computationally intensive problems, algorithms need to run in parallel.

Within the last few years, researchers have begun to recognize the need to
develop parallel implementations of PSO, publishing many papers on the subject.
The methods they have used include various synchronous algorithms
\cite{belal-ijicis04,chu-sci06,jin-aps05,parsopoulos-aia04,schutte-ijnme04},
asynchronous algorithms \cite{koh-ijnme06,mostaghim-report06,venter-wcsmo05},
and algorithms based on Google's MapReduce framework \cite{mcnabb-cec07}.
Parallelizing the evaluation of the objective function can also be done in some
cases, though that is not an adaption of the PSO algorithm itself and thus is
not the focus of this paper.

All of these parallel techniques use additional processors to increase the
number of particles in the swarm, either by adding individual particles or by
adding entire new subswarms.  The number of iterations of PSO that the
algorithms can perform is thus inherently limited by the time it takes to
evaluate the objective function---additional processors add more particles, not
more iterations.  However, there are many functions for which performing more
iterations with fewer particles is a better use of resources than performing
fewer iterations with more particles.

To solve this problem, we take inspiration from speculative execution
techniques commonly used in processors.  Modern processors, when faced with a
branch on which they must wait (e.g. a memory cache miss), guess which way the
branch will go and start executing.  If the processor guessed right, execution
is much farther ahead than if it had just idly waited on the branch.  If it
guessed wrong, execution restarts right where it would have been anyway.  This
is called speculative execution or branch prediction.

The idea of speculative execution can be applied to some types of optimization
algorithms.  When the next sampling position of the algorithm does not depend
explicitly on the value of the objective function at the current position,
future positions can be speculated and evaluated in parallel with the current
position.  This could not be done in gradient methods, for instance, where the
step size for sampling depends on the gradient at the current position, but it
has been done in simulated annealing~\cite{witte-tpds91}.

Speculative evaluation is possible in PSO because the motion of particles
depends only on the \emph{positions} of the particle and its best neighbor, not
the \emph{value} of the objective function at those positions.  We do not have
to wait for a particle's function evaluation to complete in order to know where
it might move to next, so we can compute the set of all possible next positions
and send them to be evaluated along with the particle's current position.  Once
the current position of each particle has been evaluated, we can determine
which of the speculatively evaluated positions matches the branch taken by the
particle.  We have thus evaluated two iterations at once.  This can easily get
unwieldy if the set of next possible positions for each particle is large, but
a wise choice of topology limits that set to a reasonable size.

The original PSO can be reproduced exactly, two iterations at a time, using
this technique.  However, in strictly using speculative evaluation to reproduce
the original PSO, a lot of positions are speculatively evaluated whose values
get ignored after deciding which branch was the correct one.  A better use of
those speculative evaluations may be to incorporate the information into the
swarm, no longer exactly reproducing PSO but in some instances producing better
results.

\sec{pso} describes the particle swarm optimization algorithm.  \sec{sepso}
then describes how speculative evaluation can be done in parallel PSO to
perform two iterations at once.  In \sec{extra}, we discuss a variant of the
speculative algorithm which uses all of the speculative evaluations performed
instead of throwing away all but one for each particle.  We then show in
\sec{results} a comparison between the original PSO algorithm, our speculative
algorithm, and its variant.  In \sec{conclusion} and \sec{future} we conclude
and discuss future work.

@@

Particle swarm optimization was proposed in 1995 by James Kennedy and Russell
Eberhart~\cite{kennedy-icnn95}.  The algorithm is used to intelligently search
a multi-dimensional space by mimicking the swarming and flocking behavior of
birds and other animals. It is a social algorithm that depends on interaction
between particles to quickly and consistently find the optimal solution to a
given objective function.

@@

At each iteration of the algorithm, the position $\ppos_t$ and velocity
$\pvel_t$ of each particle are updated as follows:
\begin{align}
\label{eq:velupdate}
	\pvel_{t+1} &=
		\constriction \left[ \pvel_t +
			\coeff_1\URand()\otimes(\pbest - \ppos_t) +
			\coeff_2\URand()\otimes(\nbest - \ppos_t)
		\right] \\
\label{eq:posupdate}
	\ppos_{t+1} &= \ppos_t + \pvel_{t+1}
\end{align}
where \( \URand() \) is a vector of random numbers drawn from a uniform
distribution, the \( \otimes \) operator is an element-wise vector
multiplication, $\pbest$ (called personal best) is the best position the
current particle has seen, and $\nbest$ (called neighborhood best) is the best
position the neighbors of the current particle have seen.  \( \coeff_1 \), \(
\coeff_2 \), and \( \constriction \) are parameters with prescribed values
required to ensure convergence (2.05, 2.05, and .73,
respectively)~\cite{clerc-tec02}.

There are many ways of defining the neighbors of a particle.  Different
neighborhood topologies have varying performance on different kinds of
functions, as would be expected by the No Free Lunch
theorem~\cite{wolpert-tec97}.  An in-depth discussion of topologies in PSO is
beyond the scope of this paper.  The two most common topologies used in the
literature are the Ring topology and the Complete
topology~\cite{bratton-sis07}.  In this paper we use those topologies, as well
as a parallel adaptation of the Complete topology called Random that has been
shown to simulate the behavior of Complete with far less
communication~\cite{mcnabb-cec09}.  Both the Ring and the Random topologies can
be defined as having $k$ neighbors; we use 2 neighbors for all of our
experiments.

\section{Speculative Evaluation in PSO}
\label{sec:sepso}
\subsection{Overview}

The PSO algorithm just described can be parallelized by having each additional
processor add a particle or a set of particles to the swarm.  But for some
functions, a more desirable use of resources is to speed up the algorithm,
performing more iterations instead of adding more particles.  Barring
parallelizing the evaluation of the objective function itself, which may not be
possible, previous techniques have not addressed this issue.

In order to perform additional iterations by adding additional processors, we
take inspiration from speculative execution in processors.  If we can guess
what a particle's next position will be from its current position, we can
evaluate both in parallel and perform two iterations of PSO at once.  Lacking
a good way to guess a particle's next position, we can at least enumerate all
possible next positions of the particle, evaluate all of them at once, and one
of them will be right.

While in some algorithms enumerating all possible next sampling positions would
be impossible or extremely difficult, in PSO all we need to know are the
current positions of the particle and its neighbors, from which we can easily
enumerate and evaluate all possible positions of that particle at the next
iteration.  The defining components of a particle's motion
(see~\eqref{eq:velupdate}) are $\pvel_t$, $\ppos_t$, $\pbest$, and $\nbest$.
For a particle at iteration $t-1$, $\pvel_t$ and $\ppos_t$ are defined by
Equations~\eqref{eq:velupdate} and \eqref{eq:posupdate}.  The only remaining
components are $\pbest$ and $\nbest$, each of which has only a few
possibilities.  The particle already knows its personal best $\pbest$ at
iteration $t-1$, but it could be updated to $\ppos_t$ if its position at
iteration $t$ has a better value than its current personal best.  Likewise, the
particle knows its neighborhood best $\nbest$ at iteration $t-1$, though it
could be updated by any of its neighbors' positions at iteration $t$.

To enumerate all of the possible next positions of a particle, one must simply
evaluate the motion equations for all possible combinations of personal best
and neighborhood best updates that a particle could see.  This produces
$2(n+1)$ speculative positions to be evaluated, where $n$ is the number of
neighbors the particle has.  If a particle is also its own neighbor, as is most
often the case, one of the $2(n+1)$ speculative positions can be eliminated, as
a particle cannot be the source of its $\nbest$ update while not updating its
$\pbest$.  Thus we have $2(n+1)-1$ evaluations.  Table~\ref{tab:evals} lists
all possible combinations of updates that can be seen by a particle with two
neighbors other than itself, which is the case in all of the topologies we use.
If all of these speculative positions are evaluated in parallel along with the
original, non-speculative position of the particle at time $t-1$, then one of
the speculative positions has to be the correct one, and two iterations have
been evaluated at once.

\begin{table}
  \caption{All possible updates for a particle with two neighbors.}
  \label{tab:evals}
  \centering
  \begin{tabular}{cc}
	$\pbest$ updated&Source of $\nbest$ update\\
	\hline
	\hline
	No update&No update\\
	\hline
	No update&Neighbor 1\\
	\hline
	No update&Neighbor 2\\
	\hline
	Updated&No update\\
	\hline
	Updated&Neighbor 1\\
	\hline
	Updated&Neighbor 2\\
	\hline
	Updated&The particle itself\\
	\hline
  \end{tabular}
\end{table}

Because $2(n+1)-1$ speculative evaluations must be performed for each particle,
the choice of topology is an important one.  The use of the Complete topology,
where every particle is a neighbor to every other particle, would require
$O(p^2)$ speculative evaluations per iteration, where $p$ is the swarm size.
Clearly it is much more desirable to have a sparse topology, where $O(np)$ is
much smaller than $O(p^2)$.  However, some functions are better optimized with
the Complete topology and the quick spread of information it entails than with
sparse topologies.  Accordingly, we use the Random topology described
in~\cite{mcnabb-cec09}, which has been shown to effectively simulate the
Complete topology.  In that topology, 2 neighbors are chosen at random from the
swarm for each particle at each iteration.  Because a particle is also always
its own neighbor, $n$ is $3$ in that case.  In~\sec{results} we report the
results for performing speculative evaluation using both the Ring topology
(where $n$ again is $3$) and the Random topology on a number of common
benchmark functions.

It is not trivial in some parallel architectures to determine which speculative
position was the correct next position of each particle.  First we discuss the
relatively easy case of a centralized parallel PSO algorithm.  In such an
architecture, the master keeps track of all necessary information with only
trivial message passing needed.  Then we discuss the more complicated case of a
distributed algorithm, where each particle is on its own and needs to send and
receive messages to and from other particles.  Finally we discuss the further
complications of a dynamic topology, where a particle's neighbors change from
one iteration to another.

@@

A particle at iteration $t-1$ that has been moved to iteration $t$, but whose
position has not yet been evaluated, is denoted as $\noeval{\p}_t$.  Once its
position has been evaluated, but it has still not yet received information from
its neighbors, it is denoted as $\nonbest{\p}_t$.  Only when the particle has
updated its neighborhood best is it a complete particle at iteration $t$.  It is
then simply denoted as $\p_t$.

@@

In a distributed parallel PSO algorithm, individual processors not only perform
evaluations of particles, but also their movement.  The information for each
particle is not held by a central machine that directs the algorithm; instead,
each processor has the information for the particle or particles that it is in
charge of and must perform the steps of the algorithm for those particles.
Messages such as values and positions for the neighborhood best are sent
between processors.  There may still be some machine that collects information
from all of the particles and outputs the result of the algorithm, though that
machine's importance is much less than in centralized algorithms.

>>>>

\documentclass{llncs}

@@

%% PSO Stuff
\DeclareMathOperator{\URand}{U}
\DeclareMathOperator*{\argmin}{arg\;min}
\DeclareMathOperator*{\argmax}{arg\;max}
\DeclareMathOperator*{\arginf}{arg\;inf}
\DeclareMathOperator*{\argsup}{arg\;sup}
\providecommand{\pers}{\ensuremath{P}}
\providecommand{\neigh}{\ensuremath{N}}
\providecommand{\leftind}{\ensuremath{L}}
\providecommand{\rightind}{\ensuremath{R}}
\providecommand{\nURand}{\ensuremath{U^\neigh}}
\providecommand{\pURand}{\ensuremath{U^\pers}}
\providecommand{\ppos}{\ensuremath{\Vec{x}}}
\providecommand{\fppos}{\ensuremath{\Vec{fx}}}
\providecommand{\pvel}{\ensuremath{\Vec{v}}}
\providecommand{\nbest}{\ensuremath{\Vec{x}^\neigh}}
\providecommand{\fnbest}{\ensuremath{\Vec{fx}^\neigh}}
\providecommand{\pbest}{\ensuremath{\Vec{x}^\pers}}
\providecommand{\fpbest}{\ensuremath{\Vec{fx}^\pers}}
\providecommand{\constriction}{\ensuremath{\chi}}
\providecommand{\ncoeff}{\ensuremath{\phi^\neigh}}
\providecommand{\pcoeff}{\ensuremath{\phi^\pers}}
\providecommand{\obs}{\ensuremath{\Vec{\xi}}}
\providecommand{\ofunc}{\ensuremath{f}}
\providecommand{\swarm}{\ensuremath{swarm}}

%SpecExPSO Stuff
\providecommand{\indic}{\ensuremath{I}}
\providecommand{\specvel}{\ensuremath{\vec{V}}}
\providecommand{\specpos}{\ensuremath{\vec{X}}}
\providecommand{\leftn}{\ensuremath{\Vec{x}^\leftind}}
\providecommand{\rightn}{\ensuremath{\Vec{x}^\rightind}}
\providecommand{\caseset}{\ensuremath{\mathcal{C}}}
\providecommand{\casegen}{\ensuremath{c}}
\providecommand{\casedef}{\ensuremath{(\pbest,\nbest)}}
\providecommand{\casexn}{\ensuremath{(S,-)}}
\providecommand{\casexx}{\ensuremath{(S,S)}}
\providecommand{\casexl}{\ensuremath{(S,\leftind)}}
\providecommand{\casexr}{\ensuremath{(S,\rightind)}}
\providecommand{\casepn}{\ensuremath{(-,-)}}
\providecommand{\casepl}{\ensuremath{(-,\leftind)}}
\providecommand{\casepr}{\ensuremath{(-,\rightind)}}
\providecommand{\casepN}{\ensuremath{(-,N)}}
\providecommand{\casexN}{\ensuremath{(S,N)}}
\providecommand{\noeval}[1]{\ensuremath{#1^{-e}}}
\providecommand{\nonbest}[1]{\ensuremath{#1^{-n}}}
\providecommand{\p}{\ensuremath{p}}
\providecommand{\pset}{\ensuremath{\mathbf{p}}}
\providecommand{\s}{\ensuremath{s}}
\providecommand{\sset}{\ensuremath{\mathbf{s}}}
\providecommand{\nsset}{\ensuremath{\mathbf{ns}}}
\providecommand{\n}{\ensuremath{n}}
\providecommand{\nset}{\ensuremath{\mathbf{n}}}
\providecommand{\nnset}{\ensuremath{\mathbf{nn}}}

@@

@@

Particle swarm optimization (PSO) has been found to be a highly robust and
effective algorithm for solving many types of optimization problems.  For much
of the algorithm's history, PSO was run serially on a single machine.  However,
the world's computing power is increasingly coming from large clusters of
processors.  In order to efficiently utilize these resources for
computationally intensive problems, PSO needs to run in parallel.

Within the last few years, researchers have begun to recognize the need to
develop parallel implementations of PSO, publishing many papers on the subject.
The methods they have used include various synchronous algorithms
\cite{chu-2006-intelligent-parallel-pso, jin-2005-pso-antenna-designs,
parsopoulos-2004-parallel-vector-evaluated-pso,
schutte-2004-parallel-global-optimization-with-pso} and asynchronous algorithms
\cite{mostaghim-2006-multi-objective-pso-on-grids,
venter-2005-parallel-pso-asynchronous-evaluations}.  Parallelizing the
evaluation of the objective function can also be done in some cases, though
that is not an adaption of the PSO algorithm itself and thus is not the focus
of this paper.

These previous parallel techniques distribute the computation needed by the
particles in the swarm over the available processors.  If more processors are
available, these techniques increase the number of particles in the swarm,
either by adding individual particles or by adding entire new sub-swarms, but
in all cases the number of processors never exceeds the number of particles.
The number of iterations of PSO that the algorithms can perform is thus
inherently limited by the time it takes to evaluate the objective
function---additional processors add more particles, but do not make the
iterations go any faster.

For many functions there comes a point of diminishing returns with respect to
adding particles.  In \fig{evals-sphere} we show the value obtained after
50,000 function evaluations (not iterations) as a function of swarm size for
the well-know benchmark function Sphere (20 dimensions, reporting the average
of twenty runs).  Increasing the swarm size from 5 to 10 has a significant
effect on the value obtained.  However, increasing the swarm size from 16 to 30
makes the algorithm less efficient, that is, it reduces the progress the
algorithm makes per evaluation.  Other functions show similar trends, though
often the optimal swarm size is slightly larger.  For this reason, previous
work has recommended the use of a swarm size of 50 for
PSO~\cite{bratton-2007-defining-a-standard-for-pso}.  Thus, in at least some
cases, adding particles indefinitely will not yield an efficient
implementation.

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{evals_sphere}
  \caption{Function Sphere with various swarm sizes. Error bars show median and
  10th and 90th percentiles.}
  \label{fig:evals-sphere}
\end{figure}

In this paper we consider PSO parallelization strategies for clusters of
thousands of processors and functions for which a single evaluation will take
at least tens of seconds, but probably minutes or perhaps hours.  Our purpose
is to explore the question of what to do with a thousand processors when 50 or
100 particles is the recommended swarm size, and simply increasing the swarm
size encounters diminishing returns.

To solve this problem, we propose a method for performing two iterations of PSO
at the same time in parallel that we call speculative evaluation.  The name
comes from an analogy to speculative execution (also known as branch
prediction) techniques commonly used in processors.  Modern processors, when
faced with a branch on which they must wait (e.g. a memory cache miss), guess
which way the branch will go and start executing, ensuring that any changes
can be undone.  If the processor guesses right, execution is
much farther ahead than if it had idly waited on the memory reference.  If
it guesses wrong, execution restarts where it would have been anyway.

In this paper we show that the results of standard PSO can be reproduced
\emph{exactly}, two iterations at a time, using a speculative approach adapted
from speculative execution. We prove that the standard PSO equation can be
factored such that a set of speculative positions can be found which will
\emph{always} include the position computed in the next iteration.  By
computing the value of the objective function for each of the speculative
positions at the same time the algorithm evaluates the objective function for
the current position, it is possible to know the objective function values for
both the current and the next iteration at the same time.  We demonstrate this
principle by implementation and show that it produces exactly the same results
as standard PSO, but two iterations at a time.  The resulting implementation
runs efficiently on large clusters where the number of processors is much
larger than a typical or reasonable number of particles, producing better
results in less ``wall-clock'' time.

The balance of this paper is organized as follows. \sec{pso} describes the
particle swarm optimization algorithm.  \sec{sepso} describes how speculative
evaluation can be done in parallel PSO to perform two iterations at once, then
presents a comparison of the speculative algorithm with the original PSO.  In
\sec{conclusion} and \sec{future} we conclude and discuss future work.

@@

Particle swarm optimization was proposed in 1995 by James Kennedy and Russell
Eberhart~\cite{kennedy-1995-particle-swarm-optimization}.  The algorithm is
used to intelligently search a multi-dimensional space by mimicking the
swarming and flocking behavior of birds and other animals. It is a social
algorithm that depends on interaction between particles to quickly and
consistently approximate the optimal solution to a given objective function.

@@

At each iteration of the algorithm, the position $\ppos_t$ and velocity
$\pvel_t$ of each particle are updated as follows:
\begin{align}
\nonumber
	\pvel_{t+1} &=
		\constriction \bigl[ \pvel_t
			+ \pcoeff\pURand_{t}\otimes(\pbest_{t} - \ppos_{t}) \\
\label{eq:velupdate}
			& \quad \quad \quad \, + \ncoeff\nURand_{t}\otimes(\nbest_{t} - \ppos_{t})
		\bigr] \\
\label{eq:posupdate}
	\ppos_{t+1} &= \ppos_{t} + \pvel_{t+1}
\end{align}
where \( \pURand_{t} \) and \( \nURand_{t} \) are vectors of independent random
numbers drawn from a standard uniform distribution, the \( \otimes \) operator
is an element-wise vector multiplication, $\pbest$ (called personal best) is
the best position the current particle has seen, and $\nbest$ (called
neighborhood best) is the best position the neighbors of the current particle
have seen~\cite{bratton-2007-defining-a-standard-for-pso}.  The parameters \(
\ncoeff \), \( \pcoeff \), and \( \constriction \) are given prescribed values
required to ensure convergence (2.05, 2.05, and .73,
respectively)~\cite{clerc-2002-constricted-pso}.

Changing the way neighbors are defined (often called the ``topology'' in the
literature) has a significant effect on the performance of the algorithm.  In
the Ring topology, each particle has one neighbor to either side of it; in the
Complete topology, every particle is a neighbor to every other
particle~\cite{bratton-2007-defining-a-standard-for-pso}.  In all topologies a
particle is also a neighbor to itself in that its own position and value are
considered when updating the particle's neighborhood best, $\nbest$.  Thus with
$p$ particles, using the Ring topology each particle with index $i$ has three
neighbors: $i-1$, $i$ (itself), and $i+1$.  With the Complete topology, each
particle has $p$ neighbors.

In this paper we use these topologies as well as a parallel adaptation of the
Complete topology, called Random, that has been shown to approximate the
behavior of Complete with far less
communication~\cite{mcnabb-2009-large-particle-swarms}.  In the Random
topology, each particle randomly picks two other particles to share information
with at each iteration, along with itself.  Thus in both the Ring and the
Random topologies, all particles have three neighbors.

\section{Speculative Evaluation in PSO}
\label{sec:sepso}

The PSO algorithm can be trivially parallelized by distributing the processing
needed for each particle on across processors.  But as we have seen in
\fig{evals-sphere}, for some functions, and for large numbers of
processors, just adding particles reaches a point of diminishing returns.
That is, adding processors does not help us reach any given level of fitness
appreciably faster.  Instead of adding particles we employ a speculative
evaluation approach that allows us to perform iterations two at a time.

To do so we must first refactor PSO such that the determination of the value
of the objective function is separate from the rest of the computation.  For
simplicity, this discussion will describe the case where PSO is performing
function minimization using the Ring topology.  In this example, each particle
has two neighbors, the ``right neighbor'' and ``left neighbor,'' whose
positions are represented as $\rightn$ and $\leftn$ respectively.  Though we
will only describe the case of the Ring topology, this method can easily be
extended to arbitrary topologies.

The refactoring hinges on the idea that once the random coefficients are
determined, there are only a few possibile updates to $\nbest$ and $\pbest$.
For the Ring topology there are 7 possible update cases, identified in
Table~\ref{tab:evals}.  We label each case with an identifier referring to the
source of the update: a minus sign ($-$) represents no update, $L$ represents
an update to $\nbest$ coming from the left neighbor, $R$ represents an update
to $\nbest$ coming from the right neighbor, and $S$ represents an update to
either $\pbest$ or $\nbest$ coming from the particle itself.  As an example,
$\casexn$ refers to the case that the particle finds a new personal best, but
neither it nor its neighbors found a position that updated its neighborhood
best.  In the equations that follow, we refer to an unspecified update case as
$\casegen$, and to the set of cases collectively as $\caseset$.

\begin{table}
  \caption{All possible updates for a particle with two neighbors}
  \label{tab:evals}
  \centering
  \begin{tabular}{lcc}
	Identifier&Source of $\pbest$ update&Source of $\nbest$ update\\
	\hline
	\hline
	$\casepn$&No update&No update\\
	\hline
	$\casepl$&No update&Left Neighbor\\
	\hline
	$\casepr$&No update&Right Neighbor\\
	\hline
	$\casexn$&Self&No update\\
	\hline
	$\casexl$&Self&Left Neighbor\\
	\hline
	$\casexr$&Self&Right Neighbor\\
	\hline
	$\casexx$&Self&Self\\
	\hline
  \end{tabular}
\end{table}

In order to incorporate the determination of which case occurs into the
position and velocity update equations, we introduce an indicator function
$\indic_{t+1}^{\casegen}$ for each case $\casegen \in \caseset$.  When
$\casegen$ corresponds to the case actually taken by PSO,
$\indic_{t+1}^{\casegen}$ evaluates to 1; otherwise it evaluates to 0.  We can
then sum over all of the cases, and the indicator function will make all of the
terms drop to zero except for the case that actually occurs.  For example, the
indicator function for the specific case $\casexn$ can be written as follows:

\begin{align}
  \nonumber
	\indic_{t+1}^{\casexn} & (\ofunc ( \ppos_{t} ) ,\ofunc(\leftn_{t}),
	\ofunc(\rightn_{t}) ,\ofunc(\pbest_{t-1}) ,\ofunc(\nbest_{t-1}))= \\
  \label{eq:deficasexn}
	&\begin{cases}
	   1 & \text{if} \ \ofunc(\ppos_{t}) < \ofunc(\pbest_{t-1}) \\
	   &\quad \wedge \quad \ofunc(\nbest_{t-1}) < \ofunc(\ppos_{t}) \\
	   &\quad \wedge \quad \ofunc(\nbest_{t-1}) < \ofunc(\leftn_{t}) \\
	   &\quad \wedge \quad \ofunc(\nbest_{t-1}) < \ofunc(\rightn_{t}) \\
	   0 & \text{otherwise}
	\end{cases}
\end{align}

For each case $\casegen \in \caseset$, there is also a corresponding
velocity update function $\specvel_{t+1}^{\casegen}$.  When the case is
known, the specific values of $\pbest_t$ and $\nbest_t$ may be substituted
directly into~\eqref{eq:velupdate}.  For example, in case $\casexn$,
$\pbest_{t}=\ppos_{t}$, as \pbest was updated by the particle's current
position, and $\nbest_{t}=\nbest_{t-1}$, as $\nbest$ was not updated at
iteration $t$:
\begin{align}
\nonumber
	\specvel_{t+1}^{\casexn} & (\pvel_t, \ppos_{t}, \leftn_{t}, \rightn_{t},
	\pbest_{t-1}, \nbest_{t-1}, \pURand_{t}, \nURand_{t}) \\
\nonumber
		&= \constriction \bigl[ \pvel_{t} +
			\pcoeff\pURand_{t}\otimes(\ppos_{t} - \ppos_{t}) \\
\label{eq:defvcasexn}
			& \quad \quad \quad \; + \ncoeff\nURand_{t}\otimes(\nbest_{t-1} -
			\ppos_{t}) \bigr]
\end{align}

In the same way we can create notation for the position update function by
substituting into~\eqref{eq:posupdate}.  For compactness, we will drop the
parameters to $\specvel_{t+1}^{\casegen}$ since they can be inferred from the
subscripts.
\begin{align}
\label{eq:defpcasegen}
	\specpos_{t+1}^{\casegen} & (\ppos_{t}, \pvel_{t}, \leftn_{t},
	\rightn_{t} ,\pbest_{t-1} ,\nbest_{t-1}, \pURand_{t}, \nURand_{t}) \\
\nonumber
	& = \ppos_{t} + \specvel_{t+1}^{\casegen}
\end{align}

With this notation we can re-write the original PSO velocity
equation~\eqref{eq:velupdate}, introducing our sum over cases with the
indicator functions.  Again, we represent the indicator functions and velocity
functions without the parameters for compactness.  The equation becomes:
\begin{align}
\nonumber
	\pvel_{t+1} &=
		\constriction \bigl[ \pvel_t
			+ \pcoeff\pURand_{t}\otimes(\pbest_{t} - \ppos_{t}) \\
\nonumber
			& \quad \quad \quad \, + \ncoeff\nURand_{t}\otimes(\nbest_{t} -
			\ppos_{t}) \bigr] \\
\nonumber
	&= \sum_{c \in \caseset} \indic_{t+1}^{c} \ \constriction \bigl[ \pvel_t
			+ \pcoeff\pURand_{t}\otimes(\pbest_{t} - \ppos_{t}) \\
\nonumber
			& \quad \quad \quad \, + \ncoeff\nURand_{t}\otimes(\nbest_{t} -
			\ppos_{t}) \bigr]  \\
\label{eq:vel2update}
	&= \sum_{c \in \caseset} \ \indic_{t+1}^{c} \ \specvel_{t+1}^{c} 
\end{align}

Similarly, the position update equation~\eqref{eq:posupdate} becomes:
\begin{align}
\nonumber
	\ppos_{t+1} &= \ppos_{t} + \pvel_{t+1} \\
\label{eq:pos2update}
	&= \sum_{c \in \caseset} \ \indic_{t+1}^{c} \ \specpos_{t+1}^{c} 
\end{align}

The value of the objective function at $\ppos_{t+1}$ is given by:
\begin{align}
\label{eq:val2update}
	\ofunc (\ppos_{t+1}) = \sum_{c \in \caseset} \ \indic_{t+1}^{c}
	\ \ofunc(\specpos_{t+1}^{c})
\end{align}

Returning our attention to the computation of $\ppos_{t+1}$ in
\eqref{eq:pos2update} and writing it with the parameters which were omitted
above, we obtain:
\begin{align}
\nonumber
  \ppos_{t+1} &= \sum_{c \in \caseset} \\
\nonumber
	& \indic_{t+1}^{c}(\ofunc ( \ppos_{t} ) ,\ofunc(\leftn_{t}),
	\ofunc(\rightn_{t}) ,\ofunc(\pbest_{t-1}) ,\ofunc(\nbest_{t-1})) \\
\label{eq:val2updatelong}
	& \specpos_{t+1}^{c}(\ppos_{t},\pvel_{t},\leftn_{t},\rightn_{t},
	\pbest_{t-1},\nbest_{t-1},\pURand_{t}, \nURand_{t})
\end{align}

In this form the important point to notice is that there are only 7 values (for
this Ring topology) in the set $\{\specpos_{t+1}^{\casegen}: \casegen \in
\caseset\}$ and that none of them depend upon $f(\ppos_t)$ or any other
objective function evaluation at iteration $t$ (note also that while there are
random numbers in the equation, they are assumed fixed once drawn for any
particular particle at a specific iteration).  Thus PSO has been refactored
such that the algorithm can begin computing all 7 of the objective function
evaluations potentially needed in iteration $t+1$ \emph{before} $f(\ppos_t)$ is
computed.  Once the evaluation of $f(\ppos_{t})$ is completed for all particles
only one of the indicator functions $\indic_{t+1}^{\casegen}$ will be set to 1;
hence only one of the positions $\specpos_{t+1}^\casegen$ will be kept.

Although this speculative algorithm computes
$\ofunc(\specpos_{t+1}^{\casegen})$ for all $\casegen \in \caseset$, even those
for which $\indic_{t+1}^{\casegen} = 0$, these extra computations will be
ignored, and might just as well never have been computed.  We call the set of
computations $\{\ofunc(\specpos_{t+1}^{c}) : \casegen \in \caseset\}$
``speculative children'' because only one of them will be needed.

To see the value of this refactoring, suppose that 1000 processors are
available, and that the evaluation of the function to be optimized takes one
hour.  If we only want a swarm of 100 particles, 900 of the processors would be
sitting idle for an hour at every iteration, and it would take two hours to run
two iterations.  If instead we perform speculative evaluation, sending each of
the $\ofunc(\specpos_{t+1}^{c})$ to be computed at the same time as
$\ofunc(\ppos_{t})$, we would use 800 of the 1000 processors and perform two
iterations in one hour.

In order to do two iterations at once, we must use 8 times as many processors
as there are particles in the swarm.  If we were not performing speculative
evaluation, those processors might very well be used to add particles to the
swarm.  This then raises the question of whether a swarm of 100 particles doing
twice as many iterations outperforms a swarm of 800 particles.  We show in the
rest of this paper that in many, though not all, instances, a smaller swarm
performing more iterations does in fact outperform a larger swarm.

The number of speculative evaluations needed per particle depends on the number
of neighbors each particle has.  The number of update cases in a topology where
each particle has $n$ neighbors is $2(n+1)$; there are two possibilities for
updates to $\pbest$ (updated by the particle itself and not updated), and $n+1$
possibilities for updates to $\nbest$ (updated by each neighboring particle and
not updated).  When the particle is also a neighbor to itself, as is always the
case in commonly used topologies, one of the cases can be eliminated, as a
particle cannot be the source of an update to its neighborhood best while not
also updating its personal best.  Thus we have $2(n+1)-1$, or $2n+1$,
speculative evaluations per particle.  In a swarm with $p$ particles and $n$
neighbors per particle, $(2n+1)p$ speculative evaluations are needed.

Because the number of speculative evaluations depends on the number of
neighbors a particle has, the choice of topology is an important one.  The use
of the Complete topology, where every particle is a neighbor to every other
particle, would require $O(p^2)$ speculative evaluations per iteration.
Clearly it is much more desirable to have a sparse topology, where $O(np)$ is
much smaller than $O(p^2)$.  However, some functions are better optimized with
the Complete topology and the quick spread of information it entails than with
sparse topologies.  Accordingly, we use the Random topology described
in~\cite{mcnabb-2009-large-particle-swarms}, which has been shown to
effectively simulate the Complete topology.  In~\sec{results} we report the
results for performing speculative evaluation using both the Ring topology and
the Random topology on a number of common benchmark functions.

It is not trivial in some parallel architectures to determine which speculative
position was the correct next position of each particle.  In the rest of this
section we move from the mathematics of our method to its implementation.
First we discuss the relatively easy case of a centralized parallel PSO
algorithm with a master computer and many slaves.  In such an architecture, the
master keeps track of all necessary information with only trivial message
passing needed.  Then we discuss the more complicated case of a distributed
algorithm, where each particle is on its own and needs to send and receive
messages to and from other particles.  Finally we discuss the further
complications of a dynamic topology, where a particle's neighbors change from
one iteration to another.

@@

A particle at iteration $t-1$ that has been moved to iteration $t$ using
\eqref{eq:velupdate}~and~\eqref{eq:posupdate}, but whose position has
not yet been evaluated, is denoted as $\noeval{\p}_t$.  Once its position has
been evaluated, but it has still not yet received information from its
neighbors, it is denoted as $\nonbest{\p}_t$.  Only when the particle has
updated its neighborhood best is it a complete particle at iteration $t$.  It
is then simply denoted as $\p_t$.

@@

\label{sec:distributed}

In a distributed parallel PSO algorithm, individual processors not only perform
evaluations of particles, but also their movement.  The information for each
particle is not held by a central machine that directs the algorithm; instead,
each processor has the information for the particle or particles that it is in
charge of and must perform the steps of the algorithm for those
particles~\cite{mcnabb-2007-parallel-pso-using-mapreduce}.  Messages such as
values and positions for the neighborhood best are sent between processors.
There may still be some machine that collects information from all of the
particles and outputs the result of the algorithm, though that machine's
importance is much less than in centralized algorithms.
